\texttt{vqtl}: An \texttt{R} package for Mean-Variance QTL Mapping

### 2018-07-12 revision/response.tex

\EditorQuestion{
   For instance, one of the other papers could provide the benchmark data that is required for [the software] manuscript.
   So the software paper can include those files and refer to the other manuscript for details.
}
\Response{
  We have incorporated this recommendation, using the dataset from Kumar et al. 2013 that was used in one of the case studies in the manuscript titled ``Mean-Variance QTL Mapping Identifies Novel QTL for Circadian Activity and Exploratory Behavior in Mice''.
  The time required to analyze this dataset, according to number of computer cores available and number of permutations desired, is described in the ``Performance Benchmarks'' section of the software paper and the relevant files are included in its Zenodo repository.
}

\subsection*{Issues Specific to this Manuscript}

\EditorQuestion{
  As suggested by Reviewer 1 it will be very useful to have a benchmark study with times/memory requirements as a function of number of markers and sample size.
  This study will give those interested on using the package an understanding of how this package would scale.
  It will be important to include in the discussion also elements related to scalability: is the package suitable for very large-p/large-n? 
}
\Response{
  We have added this information in the ``Performance Benchmarks'' section. 
  See detailed response to Reviewer 1 comment 3. 
  Note that this has lengthened the paper somewhat, but we can move plots to supplemental if that would be preferred.
}

\EditorQuestion{
  There is a very rich literature in the study of genetic control in environmental variance in both plant and animal breeding as well as model organisms.
  This literature should be mentioned. This comment may also apply to some of the companion papers.
  [Citation recommendations provided by the editor.]
}
\Response{
  Thank you for these suggestions. 
  This literature was foundational toward the development of the work presented in this manuscript, and we agree with the editor that it should be cited in the introduction.
  We have added a paragraph to the introduction that describes how recognizing the heritability of environmental variance is a necessary step toward vQTL and mvQTL mapping (``It has long been recognized\dots'').
}

@@

% Ibáñez N, Varona L, Sorensen D, Noguera JL. A study of heterogeneity of environmental variance for slaughter weight in pigs. Animal. 2007;2:19-26.

% Sørensen P, de los Campos G, Morgante F, Mackay TFC, Sorensen D. Genetic Control of Environmental Variation of Two Quantitative Traits of Drosophila melanogaster Revealed by Whole-Genome Sequencing. Genetics. 2015;201(2):487-497. doi:10.1534/genetics.115.180273.

\section*{Reviewer 1 Comments}

\ReviewerQuestion{
  The authors mention ``in almost all traits of interest in human health and disease''.
  The detection of QTL has also been widely used in plant and animal populations.
}
\Response{
  The reviewer is quite correct. Rather than add to this sentence, we have removed the explicit reference to health and disease, recognizing that the introduction started too broadly. Instead we have included references to heritability of variance that rely heavily on studies of model organisms, livestock and crops.
}

### 2018-07-12 Package_vqt_Manuscript.tex
edits

\keywords{heteroskedastic regression, mvQTL, variance heterogeneity, DGLM}

@@

Traditional quantitative trait locus (QTL) analyses have focused on discovering ``mean QTL'', regions of the genome where allelic variation drives heterogeneity of phenotype mean, while assuming that the residual variance, that is, the intrinsic stability or noisiness of the phenotype, is identical for every individual in the mapping population.
Recent work, however, has challenged the assumption of homogeneous residual variance and sought to identify an additional type of QTL---those that influence the extent of residual variance, sometimes termed ``variance QTL'' (vQTL) \citep{Pare2010,Ronnegard2011a,Ronnegard2012,Cao2014}.

Although detection of vQTL has started to enter the mainstream of genetic analysis \citep{Yang2012,Hulse2013,Ayroles2015,Wei2016-lt,Wang2017,Wei2017-tt}, statistical tools for this purpose remain heterogeneous.
We have developed a standardized method for QTL mapping in experimental crosses that simultanously model mean and variance effects in order to detect mQTL, vQTL and a generalization of the two that we term ``mvQTL''.
Our approach, which we term ``mean-variance QTL mapping'', can currently be applied to F2 intercross and backcross designs.
In two companion articles, we characterize this method and competitors in the setting where a background factor drives variance heterogeneity (\CortyMethodsPaper), and use it to discover new QTL from two existing data resources (\CortyReanalysisPaper).

Here, we provide a practical guide to the approach using its associated \texttt{R} package \texttt{vqtl}, which implements mean-variance QTL mapping and is inter-operable with the established \texttt{qtl} package \citep{Broman2003}.
First, to generate illustrative data, we simulate an F2 intercross and four phenotypes: one phenotype determined entirely by random noise, and one with each of the three kinds of QTL.
On each phenotype we then conduct a genome scan using standard approximations to interval mapping \citep{Lander1989a,Martinez1992}, and mean-variance QTL mapping, which includes a test for mQTL, a test for vQTL, and a test for mvQTL.
The association statistics of all four tests are initially plotted in LOD score units, with drawbacks of this plotting unit discussed; then permutation scans are used to determine empirically-adjusted $p$-values, and plotting in these units is shown to to make the results of the four tests more easily comparable.
Plots are then described that communicate effects that led to the QTL's detection, and the bootstrap is used to estimate its confidence interval.
Last, we benchmark performance, using one of the datasets examined in \CortyReanalysisPaper to report how computation time varies with marker density and number of permutations.

@@

\section{Acknowledgments}
This work was primarily funded by National Institutes of General Medical Sciences (NIGMS) grants to WV (R01-GM104125,R35-GM127000) and a National Institutes of Mental Health grant to RWC (F30-MH108265).
RWC received additional support from NIGMS grant T32-GM067553 and National Library of Medicine grant T32-LM012420.

>>>>

\keywords{heteroscedastic regression, mvQTL, variance heterogeneity, DGLM}

@@

Traditional quantitative trait locus (QTL) analyses have focused on discovering ``mean QTL'' (mQTL), regions of the genome where allelic variation drives heterogeneity of phenotype mean, while assuming that the residual variance, that is, the instrinsic stability or noisiness of the phenotype, is identical for every individual in the mapping population.
It has long been recognized, however, that the residual variance is itself heritable \WV{\citep{Falconer1965,Lynch1998}}{Do they definitely talk about this?}.
This phenomenon has been described theoretically \citep{Hill2004-uo,Hill2010}, demonstrated in inbred model organisms \WV{\citep{Sorensen2015}}{Seems a bit recent} and crops \citep{Yang2012-aw}, and exploited in livestock improvement efforts \citep{Mulder2008,Ibanez-Escriche2008-ie}.
%To help dissect the genetic basis, 
Correspondingly, several groups have proposed statistical methods for mapping QTL controling the extent of this residual variance, these sometimes termed ``variance QTL'' (vQTL) \citep{Pare2010,Ronnegard2011a,Ronnegard2012,Cao2014}\WV{\citep{Soave2017,Dumitrascu2018}}{add Englehardt, Soave?}.

% Traditional quantitative trait locus (QTL) analyses have focused on discovering ``mean QTL'', regions of the genome where allelic variation drives heterogeneity of phenotype mean, while assuming that the residual variance, that is, the intrinsic stability or noisiness of the phenotype, is identical for every individual in the mapping population.
% However, it has long been recognized that the extent of residual variation is itself a heritable trait \citep{Falconer1965,Lynch1998}.
% This phenomenon has been described theoretically \citep{Hill2004-uo,Hill2010} and demonstrated in inbred model organisms \citep{Sorensen2015,Ayroles2015} and crops \citep{Yang2012-aw,Forsberg2015} and exploited in livestock improvement efforts \citep{Mulder2008,Ibanez-Escriche2008-ie}.
% Recent work, has begun to recognize heterogeneous residual variance in mapping efforts by identifying an additional type of QTL---those that influence the extent of residual variance, sometimes termed ``variance QTL'' (vQTL) \citep{Pare2010,Ronnegard2011a,Ronnegard2012,Cao2014}.

Although detection of vQTL has started to enter the mainstream of genetic analysis \citep{Yang2012,Hulse2013,Ayroles2015,Forsberg2015,Wei2016-lt,Wang2017,Wei2017-tt}, statistical tools for this purpose remain heterogeneous.
We have developed a standardized method for QTL mapping in experimental crosses, in particular F2 intercrosses and backcrosses, that simultanously models mean and variance effects in order to detect mQTL, vQTL and a generalization of the two that we term ``mvQTL''.
Our approach, which we term ``mean-variance QTL mapping'', is based on a double generalized linear model (DGLM) \citep{Smyth1989}, following the proposed use in this context by \citet{Ronnegard2011a}.
In the first of two companion articles, we characterize the method and competitors in the setting where variance heterogeneity is driven by a background factor, such as sex, batch or housing, and show that modeling these (external) variance effects improves power to detect mQTL, vQTL and mvQTL (\CortyMethodsPaper). 
In the second companion paper, we demonstrate the approach on two existing datasets and discover new mQTL and vQTL (\CortyReanalysisPaper).

Here, we provide a practical guide to the approach using its associated \texttt{R} package \texttt{vqtl},
%
%In two companion articles, we characterize the method and competitors in the setting where variance heterogeneity is driven by a background factor (\CortyMethodsPaper),  and use it to discover new QTL from two existing data resources (\CortyReanalysisPaper).
%
%The R package \texttt{vqtl} implements mean-variance QTL mapping
which is currently suitable for F2 intercrosses and backcrosses, and is inter-operable with the well-established mean QTL-oriented package for this purpose, \texttt{qtl} \citep{Broman2003}.
%We introduce the \texttt{vqtl} package as follows.
First, to generate illustrative data, we simulate an F2 intercross and four phenotypes: one phenotype determined entirely by random noise, and one with each of the three kinds of QTL.
On each phenotype we then conduct a genome scan using standard approximations to interval mapping \citep{Lander1989a,Martinez1992}, and mean-variance QTL mapping, which includes a test for mQTL, a test for vQTL, and a test for mvQTL.
The association statistics of all four tests are initially plotted in LOD score units, with drawbacks of this plotting unit discussed; then permutation scans are used to determine empirically-adjusted $p$-values, and plotting in these units is shown to to make the results of the four tests more easily comparable.
Plots are then described that communicate effects that led to the QTL's detection, and the bootstrap is used to estimate its confidence interval.
Last, we benchmark performance, using one of the datasets examined in \CortyReanalysisPaper to report how computation time varies with marker density and number of permutations.

@@

\section{Acknowledgments}
This work was funded by National Institutes of General Medical Sciences grants R01-GM104125 (RWC,WV), R35-GM127000 (RWC,WV), T32-GM067553 (RWC); National Heart, Lung and Blood Institute grant R21 HL126045 (RWC,WV); National Library of Medicine grant T32-LM012420 (RWC); and a National Institute of Mental Health grant F30-MH108265 (RWC).
